\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}

### 2013-11-26 docs/paper/paper.tex
twee scriptjes toegevoegd om gemiddelden van permutaties uit te rekenen en om implay met een colormap aan te roepen (automatisch een groter scherm is nog niet gelukt)

@@

Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
A causal relation between variables $X$ and $Y$ is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
An ancestor is an (possibly indirect) cause of its descendant.
A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
Image {plaatje!} shows an example of a Bayesian network.
A causal discovery method attempts at finding a real-world Bayesian network.
But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
As such, we represent our real-world, underlying Bayesian network with a much simpler model.
Such a model contains a reduced number of variables to be able to analyse and visualize the data.

@@

We define a trail as a set of edges which would make up a path if those edges would be undirected.
We can now define directional separation, or d-separation.
Two nodes $X$ and $Y$ are d-separated given a set of nodes $Z$ (with $X, Y \notin Z$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
\begin{itemize}
\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $Z$.
\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $Z$.
\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $Z$.
\end{itemize}
If $X$ and $Y$ are d-separated given $Z$, $Z$ is called a separating set of $X$ and $Y$.

\subsection{The PC-algorithm}
Whether there is a dependency between two variables can be concluded by using a statistical test.
In this way, a graph can be inferred representing all dependencies between nodes.
However, this only results in an undirected graph, as it does not give any information about causality.
To derive causal relations, we can make use of conditional dependence.
A set of rules for finding a causal relation $X \rightarrow Y$ are used in the PC-algorithm.

One of those rules is based on conditional dependency given a set variables other than $X$ and $Y$, and .

\subsection{PC Algorithm}

>>>>

% ### dots at the end of arrows and lines ###

\def \orightarrow {\circ\hspace{-0.42em}\rightarrow}
\def \oleftarrow {\leftarrow\hspace{-0.42em}\circ}
\def \orightline {\circ\hspace{-0.16em}-}
\def \oleftline {-\hspace{-0.16em}\circ}
\def \oline {\circ\hspace{-0.38em}-\hspace{-0.4em}\circ}
\def \srightarrow {\text{\textasteriskcentered}\hspace{-0.62em}\rightarrow}
\def \sleftarrow {\leftarrow\hspace{-0.62em}\text{\textasteriskcentered}}
\def \srightline {\text{\textasteriskcentered}\hspace{-0.16em}-}
\def \sleftline {-\hspace{-0.16em}\text{\textasteriskcentered}}
\def \sline {\text{\textasteriskcentered}\hspace{-0.38em}-\hspace{-0.4em}\text{\textasteriskcentered}}

@@

Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
If there is an edge between two variables $X$ and $Y$, this is denoted as $X - Y$.
A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
An ancestor is an (possibly indirect) cause of its descendant.
A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
Image {plaatje!} shows an example of a Bayesian network.
A causal discovery method attempts at finding a real-world Bayesian network.
But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
As such, we represent our real-world, underlying Bayesian network with a much simpler model.
Such a model contains a reduced number of variables to be able to analyse and visualize the data.

@@

We define a trail as a set of edges which would make up a path if those edges would be undirected.
We can now define directional separation, or d-separation.
Two nodes $X$ and $Y$ are d-separated given a set of nodes $S$ (with $X, Y \notin S$) if there is a trail from $X$ to $Y$ for which at least one of the following holds:
\begin{itemize}
\item The trail contains a structure $a \rightarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \leftarrow b \rightarrow c$ such that $b$ is in $S$.
\item The trail contains a structure $a \rightarrow b \leftarrow c$ (in which $b$ is called a collider) such that $b$ and its descendants are not in $S$.
\end{itemize}
If $X$ and $Y$ are d-separated given $S$, $S$ is called a separating set of $X$ and $Y$. Every two independent variables have a separating set; if they are unconditionally independent, his is the empty set.

\subsection{The PC-algorithm}
$\srightarrow$
$\sleftarrow$
$\srightline$
$\sleftline$
$\sline$
\cite{?}
Whether there is a dependency between two variables can be concluded by using a statistical test.
In this way, a graph can be inferred representing all dependencies between nodes.
However, this only results in an undirected graph, as it does not give any information about causality.
A set of rules for finding a causal relation between nodes, using the undirected graph, are used in the PC-algorithm.

One of these rules attempts to find so-called v-structures, which are of the form $X \rightarrow Z \leftarrow Y$.
V-structures may be found for every three nodes $X$, $Y$ and $Z$ for which $X \sline Z$, $Y \sline Z$ and in which $X$ and $Y$ are (possibly conditially) independent, .
As $X$ and $Y$ are independent, they have a separating set $S$ (if they are unconditially independent, $S$ is the empty set).
If $Z$ is not in $S$, then the only possible causal relation can be that of a v-structure, so the edges can be oriented as $X \srightarrow Z \sleftarrow Y$.

The other rules are based on arrows that have already been found.
\begin{itemize}
\item If there is an edge $X \sline Y$, and if there is a directed path from $X$ to $Y$, then $X \sline Y$ can be oriented as $X \srightarrow Y$.
\item If there are edges $X \srightarrow Y$ and $Y \sleftline Z$, if the latter is not an edge $Y \sleftarrow Z$, and if $X$ and $Z$ are not connected, orient $Y \sline Z$ as $Y \rightarrow Z$. 
\end{itemize}
